---
title: 'Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)'
author: "Blair Johnson"
output:
  pdf_document: default
  word_document: default
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
library(tidyverse)
library(agricolae)
library(dplyr)
library(lubridate)
library(ggplot2)
NTL.lake.data<-read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

NTL.lake.data$sampledate <-as.Date(NTL.lake.data$sampledate, format="%m/%d/%y")

#2

mytheme <- theme_bw(base_size=12) +
  theme(axis.text = element_text(color="black"),
        legend.position = "bottom")
theme_set(mytheme)


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July does not change with depth across all lakes
Ha: Mean lake temperature recorded during July changes with depth across all lakes


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

NTL.data.wrangled <- 
  NTL.lake.data %>%
   select(lakename, year4, daynum, depth, temperature_C, sampledate) %>%
   filter(!is.na(temperature_C)) %>%
  mutate(month=month(sampledate)) %>%
  filter(month==7) %>%
  select(lakename, year4, daynum, depth, temperature_C)

  
#5

NTL.data.scatterplot <-
  ggplot(NTL.data.wrangled, aes(x=depth, y=temperature_C, color=depth)) + geom_point() + 
  geom_smooth(method=lm) +ylim(0,35) +xlab("Depth") +ylab("Temperature in Celsius")
print(NTL.data.scatterplot)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: This figure shows that temperature decreases as the depth increases. Therefore, there is a negative correlation between the two variables. The distribution of points also show that there are more samples at depths below 10.  


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

lm.NTL.data <- lm(data=NTL.data.wrangled, temperature_C ~ depth)
summary(lm.NTL.data)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: This model shows that there is a negative correlation between lake temperature and temperature depth. For every 1m change in depth, the temperature decreases by -1.94621. With an R-squared of 0.7387, this shows that 73.87% 
The degrees of freedom is 9726 which indicates that there are 9726 logically independent variables in the dataset. The degrees of freedom is the number of observations we have to estimate. Since there are 9728 observations and a 9726 degrees of freedom, we have two parameters in the dataset. The p-value is low (<2e-16) which shows that this result is statistically significant.  

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

NTL.AIC.1 <- lm (data=NTL.data.wrangled, temperature_C ~ year4 + daynum + depth) 
step(NTL.AIC.1)

NTL.AIC.2 <- lm (data=NTL.data.wrangled, temperature_C ~ daynum + depth)
step(NTL.AIC.2)

NTL.AIC.3 <- lm (data=NTL.data.wrangled, temperature_C ~ year4 + depth)
step(NTL.AIC.3)

NTL.AIC.4 <- lm (data=NTL.data.wrangled, temperature_C ~ year4 + daynum)
step(NTL.AIC.4)

#NTL.AIC.1 with year4, daynum, and depth is best suited

#10

NTL.regression <-lm(data=NTL.data.wrangled, temperature_C ~ year4 + daynum + depth)
summary(NTL.regression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: When using the AIC method, we see that the final set of explanatory variables are year4, daynum, and date. We use these variables because the AIC is the lowest relative to other AIC variable combinations. When running a linear regression model, we see that this model explains 74.12% (adjusted R-squared of 0.7411) of the observed variance. This model is a slight improvement over the model only using depth where the R-squared is 0.7387 but the improvement is not very significant. 

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

NTL.anova <- aov(data=NTL.data.wrangled, temperature_C ~ lakename)
summary(NTL.anova)


NTL.anova.2 <- lm(data=NTL.data.wrangled, temperature_C ~ lakename)
summary(NTL.anova.2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: With a p-value of less than 2.2e-16, we reject the null hypothesis and therefore the mean temperature among the lakes are not the same. Therefore, there is a significant difference in mean temperature.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

NTL.temp.depth <- ggplot(NTL.data.wrangled, aes(x=depth, y=temperature_C, color=lakename)) + geom_point(alpha = 0.5) + ylim(0,35) + labs(color="Lake Name", y ="Temperature", x="Depth") + geom_smooth(method = "lm", se =FALSE) +theme(legend.text = element_text(size = 10), legend.position = "right")
print(NTL.temp.depth)


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(NTL.anova)

NTL.totals.groups <-HSD.test(NTL.anova, "lakename", group = TRUE) 
NTL.totals.groups


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Based on the findings above, Ward Lake, Paul Lake, and Peter Lake have the same mean temperature statistically speaking because they all have the "c" grouping. There is no lake that has a mean temperature that is statistically different from all the other lakes. 


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: The t-test would be a good way to explore whether the lakes have distinct mean temperatures. 
